---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Norimon

<!-- badges: start -->
```{r, echo = F}
library(badger)
```
`r badge_lifecycle("experimental")` 
`r badge_devel("NINAnor/Norimon", "blue")` 
`r badge_github_actions("NINAnor/Norimon")` 
`r badge_travis("NINAnor/Norimon")` 
`r badge_code_size("NINAnor/Norimon")`

[![R-CMD-check](https://github.com/NINAnor/Norimon/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NINAnor/Norimon/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Code for the Norwegian Insect Monitoring program, by the Norwegian Institute for Nature Research

## Installation

You can install the development version of Norimon from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("NINAnor/Norimon")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r connect}
library(Norimon)
connect_to_insect_db()
```


```{r beetles}
beetles <- get_observations(subset_orders = "Coleoptera")
beetles

```
